---
title: "Cost effectiveness analysis of VRD vs RD using 3-state Markov Model"
author: "Diana Flora Namaemba"
date: "2024-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Setting up a Markov Model for the Bortezomib + Lenalidomide + Dexamethasone (VRD)  the Intervention Arm
A markov model of 3-states (Progressive Free Survival(PFS), Progressive Disease (PD), Death), setting up the transitional matrix
```{r}
#Load the Shape Package
library(shape)
#Load the markovchain Package
library(markovchain)
#Load the diagram package
library(diagram)

#Creating the transition matrix of VRD
tm_vrd <- matrix(c(0.98144,0.01856, 0, 0,0.97638, 0.02362, 0,0,1),nrow = 3, byrow = TRUE)
tm_vrd

#create the Discrete Time Markov Chain
vrd_trans <- new("markovchain",transitionMatrix=tm_vrd, states=c("PFS","PD", "Death"), name="Markov Chain for VRD") 
vrd_trans

#Visualize the Movement from one state to another
plot(vrd_trans)
```

### List of all parameters in the model
The parameters include: transitional probabilities, costs, utilities e.t.c
```{r}
#List of all parameters in VRD
params_vrd = list(
  #Transition Probability from PFS to PD
  p_pfs_pd_vrd = 0.01856, 
  #Transition Probability from PD to Death
  p_pd_death_vrd = 0.02362, 

  #Cost incurred in PFS State during induction, Maintenance and in Adverse Events, when some payments were made by NHIF(Insurance)
  NHIF_Induction_AE_PFS_vrd = 20585.32951,
  
  #Cost incurred in PD State during induction, Maintenance and in Adverse Events, when some payments were made by NHIF(Insurance)
  NHIF_Induction_AE_PD_vrd = 6719.213517,
  
  #Cost incurred in PFS State during induction, Maintenance, adverse Events and including non-medical costs, when some payments were made by NHIF(Insurance)
  NHIF_Induction_AE_nonmedicalcost_PFS_vrd = 25463.70789,
  
  #Cost incurred in PD State during induction, Maintenance, adverse Events and including non-medical costs, when some payments were made by NHIF(Insurance)
  NHIF_Induction_AE_nonmedicalcost_PD_vrd = 11305.42041,
  
  #Cost incurred in PFS State during induction, Maintenance and adverse Events, when some payments were made by Cash
  CASH_Induction_AE_PFS_vrd =19499.816,
  
  #Cost incurred in PD State during induction, Maintenance and adverse Events when some payments were made by Cash
  CASH_Induction_AE_PD_vrd = 5610.110069,
  
  #Cost incurred in PFS State during induction, Maintenance, adverse Events and including non-medical costs, when some payments were made by CASH
  CASH_Induction_AE_nonmedical_cost_PFS_vrd = 24378.19438,
  
  #Cost incurred in PD State during induction, Maintenance, adverse Events and including non-medical costs, when some payments were made by Cash
  CASH_Induction_AE_nonmedical_cost_PD_vrd = 10196.31697,
  
  #Utilities in PFS state
  U_PFS_vrd = 0.81,
  #Utilities in PD stat
  U_PD_vrd = 0.645
)
```

### A Model that calculates the QALYs and Average Cost per patient

```{r}
#Create a function of the parameters
model_vrd = function(.params_vrd) {
  with(.params_vrd, {
    #Number of cycles in the life horizon of the study
    n_cycles_vrd = 72
    #Number of States
    n_states_vrd = 3
    #Number of Patients
    n_patients_vrd = 1000
    #The states in VRD
    v_states_names_vrd = c("PFS", "PD", "Death")
    #Create a transition matrix with zero values, its 3 by 3 because of 3-states
    m_p_vrd = matrix(0, nrow = 3, ncol = 3,
                 dimnames = list(from = v_states_names_vrd,
                                 to = v_states_names_vrd))
    #Update the matrix with the actual values of moving from one state to another
    #Update the transitional probability of moving from PFS to PFS
    m_p_vrd["PFS", "PFS"] = 1 - p_pfs_pd_vrd
    #Update the transitional probability of moving from PFS to PD
    m_p_vrd["PFS", "PD"] = p_pfs_pd_vrd
    #Update the transitional probability of moving from PFS to Death
    m_p_vrd["PFS", "Death"] = 0
    #Update the transitional probability of moving from PD to Death
    m_p_vrd["PD", "Death"] = p_pd_death_vrd
    #Update the transitional probability of moving from PD to PFS
    m_p_vrd["PD", "PFS"] = 0
    #Update the transitional probability of moving from PD to PD
    m_p_vrd["PD", "PD"] = 1- p_pd_death_vrd
    #Update the transitional probability of moving from Death to Death
    m_p_vrd["Death", "Death"] = 1
    
    #Create state membership matrix
    state_membership_vrd = array(NA_real_,
                             dim = c(n_cycles_vrd, n_states_vrd),
                             dimnames = list(cycle = 1:n_cycles_vrd,
                                             state = v_states_names_vrd))
    #Update the first state membership with 1000, at the start all the patients are in PFS state
    state_membership_vrd[1, ] = c(n_patients_vrd, 0, 0)
    
    #Update the rest of the state membership with the number of patients at each state and each cycle
    for (i in 2:n_cycles_vrd) {
      state_membership_vrd[i, ] = state_membership_vrd[i-1, ] %*% m_p_vrd
    }
    
    #A zero matrix for cost incurred in Induction, Maintenance and Adverse events paid for by NHIF and QALYs incurred in different states
    m_payoffs_NHIF_Induction_AE_vrd = matrix(0, nrow = 3, ncol = 2,
                       dimnames = list(state = v_states_names_vrd,
                                       payoffs = c("Cost", "QALYs")))
    
    #Update actual cost and QALYs in PFS and PD State when cost is paid by NHIF for Induction, Maintenance and Adverse events
    m_payoffs_NHIF_Induction_AE_vrd["PFS", "Cost"] = NHIF_Induction_AE_PFS_vrd
    m_payoffs_NHIF_Induction_AE_vrd["PFS", "QALYs"] = U_PFS_vrd
    m_payoffs_NHIF_Induction_AE_vrd["PD", "Cost"] = NHIF_Induction_AE_PD_vrd
    m_payoffs_NHIF_Induction_AE_vrd["PD", "QALYs"] = U_PD_vrd
    
    #A zero matrix for cost incurred in Induction, Maintenance, Adverse events and non-medical costs and paid for by NHIF and QALYs incurred in different states
    m_payoffs_NHIF_Induction_AE_nonmedical_vrd = matrix(0, nrow = 3, ncol = 2,
                                         dimnames = list(state = v_states_names_vrd,
                                                         payoffs = c("Cost", "QALYs")))
    
    #Update actual cost and QALYs in PFS and PD State when cost is paid by NHIF for Induction, Maintenance, Adverse events and non-medical cost
    m_payoffs_NHIF_Induction_AE_nonmedical_vrd["PFS", "Cost"] = NHIF_Induction_AE_nonmedicalcost_PFS_vrd
    m_payoffs_NHIF_Induction_AE_nonmedical_vrd["PFS", "QALYs"] = U_PFS_vrd
    m_payoffs_NHIF_Induction_AE_nonmedical_vrd["PD", "Cost"] = NHIF_Induction_AE_nonmedicalcost_PD_vrd
    m_payoffs_NHIF_Induction_AE_nonmedical_vrd["PD", "QALYs"] = U_PD_vrd
    
    #A zero matrix for cost incurred in Induction, Maintenance and  Adverse events and paid for by Cash and QALYs incurred in different states
    m_payoffs_CASH_Induction_AE_vrd = matrix(0, nrow = 3, ncol = 2,
                                         dimnames = list(state = v_states_names_vrd,
                                                         payoffs = c("Cost", "QALYs")))
    
    #Update actual cost and QALYs in PFS and PD State when cost is paid by Cash for Induction, Maintenance, and Adverse events
    m_payoffs_CASH_Induction_AE_vrd["PFS", "Cost"] = CASH_Induction_AE_PFS_vrd
    m_payoffs_CASH_Induction_AE_vrd["PFS", "QALYs"] = U_PFS_vrd
    m_payoffs_CASH_Induction_AE_vrd["PD", "Cost"] = CASH_Induction_AE_PD_vrd
    m_payoffs_CASH_Induction_AE_vrd["PD", "QALYs"] = U_PD_vrd
    
    #A zero matrix for cost incurred in Induction, Maintenance, Adverse events and non-medical costs and paid for by Cash and QALYs incurred in different states
    m_payoffs_cash_Induction_AE_nonmedical_vrd = matrix(0, nrow = 3, ncol = 2,
                                                    dimnames = list(state = v_states_names_vrd,
                                                                    payoffs = c("Cost", "QALYs")))
    
    #Update actual cost and QALYs in PFS and PD State when cost is paid by Cash for Induction, Maintenance,Adverse events and non-medical costs
    m_payoffs_cash_Induction_AE_nonmedical_vrd["PFS", "Cost"] = CASH_Induction_AE_nonmedical_cost_PFS_vrd
    m_payoffs_cash_Induction_AE_nonmedical_vrd["PFS", "QALYs"] = U_PFS_vrd
    m_payoffs_cash_Induction_AE_nonmedical_vrd["PD", "Cost"] = CASH_Induction_AE_nonmedical_cost_PD_vrd
    m_payoffs_cash_Induction_AE_nonmedical_vrd["PD", "QALYs"] = U_PD_vrd
    
    #View the matrix with the updated costs and QALYs for each cost- payment option
    m_payoffs_CASH_Induction_AE_vrd
    m_payoffs_cash_Induction_AE_nonmedical_vrd
    m_payoffs_NHIF_Induction_AE_vrd
    m_payoffs_NHIF_Induction_AE_nonmedical_vrd
    
    
    #total  cost and QALYs for all patients for the different cost options
    m_payoffs_CASH_Induction_AE_trace_vrd = state_membership_vrd %*% m_payoffs_CASH_Induction_AE_vrd
    m_payoffs_cash_Induction_AE_nonmedical_trace_vrd= state_membership_vrd %*%m_payoffs_cash_Induction_AE_nonmedical_vrd
    m_payoffs_NHIF_Induction_AE_trace_vrd = state_membership_vrd %*%m_payoffs_NHIF_Induction_AE_vrd
    m_payoffs_NHIF_Induction_AE_nonmedical_trace_vrd = state_membership_vrd %*%m_payoffs_NHIF_Induction_AE_nonmedical_vrd
    
    average_cost_QALYS_vrd=list(
    #Average cost and QALYS per patients for each payment option
    colSums(m_payoffs_CASH_Induction_AE_trace_vrd) / n_patients_vrd,
    colSums(m_payoffs_cash_Induction_AE_nonmedical_trace_vrd) / n_patients_vrd,
    colSums(m_payoffs_NHIF_Induction_AE_trace_vrd) / n_patients_vrd,
    colSums(m_payoffs_NHIF_Induction_AE_nonmedical_trace_vrd) / n_patients_vrd
)
    #Adding labels to each output
    label_names <- c("Average Cost and QALYs (In Induction, Maintenance and AE paid through CASH)",
                     "Average Cost and QALYs (In Induction, Maintenance, AE and Non-Medical paid through CASH)",
                     "Average Cost and QALYs (In Induction, Maintenance and AE paid through NHIF)",
                     "Average Cost and QALYs (In Induction, Maintenance, AE and Non-Medical paid through NHIF)")
  #Title for the overall output
    cat("Summary of Payoffs Matrix (Normalized by Number of Patients):\n")
    # Print each labeled output
    for (i in seq_along(average_cost_QALYS_vrd)) {
      cat("\n[", i, "]", label_names[i], ":\n")
      print(average_cost_QALYS_vrd[[i]])
}
}
)
}
```
```{r}
#Print the Average QALYs and Cost per Patient
model_vrd(params_vrd)
```

### Setting up a Markov Model for Lenalidomide + Dexamethasone (RD)  the Reference Arm
A markov model of 3-states (Progressive Free Survival(PFS), Progressive Disease (PD), Death), setting up the transitional matrix
```{r}
#Creating the transition matrix of RD
tm_rd <- matrix(c(0.97369,0.02631, 0, 0,0.97638, 0.02362, 0,0,1),nrow = 3, byrow = TRUE)
tm_rd

#create the Discrete Time Markov Chain
rd_trans <- new("markovchain",transitionMatrix=tm_rd, states=c("PFS","PD", "Death"), name="Markov Chain for RD") 
rd_trans

#Visualize the Movement from one state to another
plot(rd_trans)
```

### List of all parameters in the model
The parameters include: transitional probabilities, costs, utilities e.t.c
```{r}
#List of all parameters in RD
params_rd = list(
  #Transition Probability from PFS to PD
  p_pfs_pd_rd = 0.02631, 
  #Transition Probability from PD to Death
  p_pd_death_rd = 0.02362, 

  #Cost incurred in PFS State during induction, Maintenance and in Adverse Events, when some payments were made by NHIF(Insurance)
  NHIF_Induction_AE_PFS_rd = 17196.47715,
  
  #Cost incurred in PD State during induction, Maintenance and in Adverse Events, when some payments were made by NHIF(Insurance)
  NHIF_Induction_AE_PD_rd = 25094.08297,
  
  #Cost incurred in PFS State during induction, Maintenance, adverse Events and including non-medical costs, when some payments were made by NHIF(Insurance)
  NHIF_Induction_AE_nonmedicalcost_PFS_rd = 22677.24638,
  
  #Cost incurred in PD State during induction, Maintenance, adverse Events and including non-medical costs, when some payments were made by NHIF(Insurance)
  NHIF_Induction_AE_nonmedicalcost_PD_rd = 30007.87607,
  
  #Cost incurred in PFS State during induction, Maintenance and adverse Events, when some payments were made by Cash
  CASH_Induction_AE_PFS_rd =16323.78485,
  
  #Cost incurred in PD State during induction, Maintenance and adverse Events when some payments were made by Cash
  CASH_Induction_AE_PD_rd = 23932.35883,
  
  #Cost incurred in PFS State during induction, Maintenance, adverse Events and including non-medical costs, when some payments were made by CASH
  CASH_Induction_AE_nonmedical_cost_PFS_rd = 21804.55408,
  
  #Cost incurred in PD State during induction, Maintenance, adverse Events and including non-medical costs, when some payments were made by Cash
  CASH_Induction_AE_nonmedical_cost_PD_rd = 28846.15193,
  
  #Utilities in PFS state
  U_PFS_rd = 0.81,
  #Utilities in PD stat
  U_PD_rd = 0.645
)
```

### A Model that calculates the QALYs and Average Cost per patient

```{r}
#Create a function of the parameters
model_rd = function(.params_rd) {
  with(.params_rd, {
    #Number of cycles in the life horizon of the study
    n_cycles_rd = 72
    #Number of States
    n_states_rd = 3
    #Number of Patients
    n_patients_rd = 1000
    #The states in VRD
    v_states_names_rd = c("PFS", "PD", "Death")
    #Create a transition matrix with zero values, its 3 by 3 because of 3-states
    m_p_rd = matrix(0, nrow = 3, ncol = 3,
                 dimnames = list(from = v_states_names_rd,
                                 to = v_states_names_rd))
    #Update the matrix with the actual values of moving from one state to another
    #Update the transitional probability of moving from PFS to PFS
    m_p_rd["PFS", "PFS"] = 1 - p_pfs_pd_rd
    #Update the transitional probability of moving from PFS to PD
    m_p_rd["PFS", "PD"] = p_pfs_pd_rd
    #Update the transitional probability of moving from PFS to Death
    m_p_rd["PFS", "Death"] = 0
    #Update the transitional probability of moving from PD to Death
    m_p_rd["PD", "Death"] = p_pd_death_rd
    #Update the transitional probability of moving from PD to PFS
    m_p_rd["PD", "PFS"] = 0
    #Update the transitional probability of moving from PD to PD
    m_p_rd["PD", "PD"] = 1- p_pd_death_rd
    #Update the transitional probability of moving from Death to Death
    m_p_rd["Death", "Death"] = 1
    
    #Create state membership matrix
    state_membership_rd = array(NA_real_,
                             dim = c(n_cycles_rd, n_states_rd),
                             dimnames = list(cycle = 1:n_cycles_rd,
                                             state = v_states_names_rd))
    #Update the first state membership with 1000, at the start all the patients are in PFS state
    state_membership_rd[1, ] = c(n_patients_rd, 0, 0)
    
    #Update the rest of the state membership with the number of patients at each state and each cycle
    for (i in 2:n_cycles_rd) {
      state_membership_rd[i, ] = state_membership_rd[i-1, ] %*% m_p_rd
    }
    
    #A zero matrix for cost incurred in Induction, Maintenance and Adverse events paid for by NHIF and QALYs incurred in different states
    m_payoffs_NHIF_Induction_AE_rd = matrix(0, nrow = 3, ncol = 2,
                       dimnames = list(state = v_states_names_rd,
                                       payoffs = c("Cost", "QALYs")))
    
    #Update actual cost and QALYs in PFS and PD State when cost is paid by NHIF for Induction, Maintenance and Adverse events
    m_payoffs_NHIF_Induction_AE_rd["PFS", "Cost"] = NHIF_Induction_AE_PFS_rd
    m_payoffs_NHIF_Induction_AE_rd["PFS", "QALYs"] = U_PFS_rd
    m_payoffs_NHIF_Induction_AE_rd["PD", "Cost"] = NHIF_Induction_AE_PD_rd
    m_payoffs_NHIF_Induction_AE_rd["PD", "QALYs"] = U_PD_rd
    
    #A zero matrix for cost incurred in Induction, Maintenance, Adverse events and non-medical costs and paid for by NHIF and QALYs incurred in different states
    m_payoffs_NHIF_Induction_AE_nonmedical_rd = matrix(0, nrow = 3, ncol = 2,
                                         dimnames = list(state = v_states_names_rd,
                                                         payoffs = c("Cost", "QALYs")))
    
    #Update actual cost and QALYs in PFS and PD State when cost is paid by NHIF for Induction, Maintenance, Adverse events and non-medical cost
    m_payoffs_NHIF_Induction_AE_nonmedical_rd["PFS", "Cost"] = NHIF_Induction_AE_nonmedicalcost_PFS_rd
    m_payoffs_NHIF_Induction_AE_nonmedical_rd["PFS", "QALYs"] = U_PFS_rd
    m_payoffs_NHIF_Induction_AE_nonmedical_rd["PD", "Cost"] = NHIF_Induction_AE_nonmedicalcost_PD_rd
    m_payoffs_NHIF_Induction_AE_nonmedical_rd["PD", "QALYs"] = U_PD_rd
    
    #A zero matrix for cost incurred in Induction, Maintenance and  Adverse events and paid for by Cash and QALYs incurred in different states
    m_payoffs_CASH_Induction_AE_rd = matrix(0, nrow = 3, ncol = 2,
                                         dimnames = list(state = v_states_names_rd,
                                                         payoffs = c("Cost", "QALYs")))
    
    #Update actual cost and QALYs in PFS and PD State when cost is paid by Cash for Induction, Maintenance, and Adverse events
    m_payoffs_CASH_Induction_AE_rd["PFS", "Cost"] = CASH_Induction_AE_PFS_rd
    m_payoffs_CASH_Induction_AE_rd["PFS", "QALYs"] = U_PFS_rd
    m_payoffs_CASH_Induction_AE_rd["PD", "Cost"] = CASH_Induction_AE_PD_rd
    m_payoffs_CASH_Induction_AE_rd["PD", "QALYs"] = U_PD_rd
    
    #A zero matrix for cost incurred in Induction, Maintenance, Adverse events and non-medical costs and paid for by Cash and QALYs incurred in different states
    m_payoffs_cash_Induction_AE_nonmedical_rd = matrix(0, nrow = 3, ncol = 2,
                                                    dimnames = list(state = v_states_names_rd,
                                                                    payoffs = c("Cost", "QALYs")))
    
    #Update actual cost and QALYs in PFS and PD State when cost is paid by Cash for Induction, Maintenance,Adverse events and non-medical costs
    m_payoffs_cash_Induction_AE_nonmedical_rd["PFS", "Cost"] = CASH_Induction_AE_nonmedical_cost_PFS_rd
    m_payoffs_cash_Induction_AE_nonmedical_rd["PFS", "QALYs"] = U_PFS_rd
    m_payoffs_cash_Induction_AE_nonmedical_rd["PD", "Cost"] = CASH_Induction_AE_nonmedical_cost_PD_rd
    m_payoffs_cash_Induction_AE_nonmedical_rd["PD", "QALYs"] = U_PD_rd
    
    #View the matrix with the updated costs and QALYs for each cost- payment option
    m_payoffs_CASH_Induction_AE_rd
    m_payoffs_cash_Induction_AE_nonmedical_rd
    m_payoffs_NHIF_Induction_AE_rd
    m_payoffs_NHIF_Induction_AE_nonmedical_rd
    
    
    #total  cost and QALYs for all patients for the different cost options
    m_payoffs_CASH_Induction_AE_trace_rd = state_membership_rd %*% m_payoffs_CASH_Induction_AE_rd
    m_payoffs_cash_Induction_AE_nonmedical_trace_rd= state_membership_rd %*%m_payoffs_cash_Induction_AE_nonmedical_rd
    m_payoffs_NHIF_Induction_AE_trace_rd = state_membership_rd %*%m_payoffs_NHIF_Induction_AE_rd
    m_payoffs_NHIF_Induction_AE_nonmedical_trace_rd = state_membership_rd %*%m_payoffs_NHIF_Induction_AE_nonmedical_rd
    
    average_cost_QALYS_rd=list(
    #Average cost and QALYS per patients for each payment option
    colSums(m_payoffs_CASH_Induction_AE_trace_rd) / n_patients_rd,
    colSums(m_payoffs_cash_Induction_AE_nonmedical_trace_rd) / n_patients_rd,
    colSums(m_payoffs_NHIF_Induction_AE_trace_rd) / n_patients_rd,
    colSums(m_payoffs_NHIF_Induction_AE_nonmedical_trace_rd) / n_patients_rd
)
    #Adding labels to each output
    label_names_rd <- c("Average Cost and QALYs (In Induction, Maintenance and AE paid through CASH)",
                     "Average Cost and QALYs (In Induction, Maintenance, AE and Non-Medical paid through CASH)",
                     "Average Cost and QALYs (In Induction, Maintenance and AE paid through NHIF)",
                     "Average Cost and QALYs (In Induction, Maintenance, AE and Non-Medical paid through NHIF)")
  #Title for the overall output
    cat("Summary of Payoffs Matrix (Normalized by Number of Patients):\n")
    # Print each labeled output
    for (i in seq_along(average_cost_QALYS_rd)) {
      cat("\n[", i, "]", label_names_rd[i], ":\n")
      print(average_cost_QALYS_rd[[i]])
}
}
)
}
```
```{r}
#Print the Average QALYs and Cost per Patient
model_rd(params_rd)
```
